A Tale of Two Transects

### 2020-02-05 writing/transect_paper.Rmd
Write positive control sequences to a file
modify results on read counts correlations to include Illumina.

(ref:readcounts) Number of Amplicon Sequence Variants (ASV) and reads at different pipeline stages.
*Raw*: raw reads as delivered by sequencing center;
*Trim*: reads with primers and demultiplexing indices removed;
*LSUx*: reads with a positive CM hit for 5.8S, allowing regions to be extracted;
*ITS2*: ASVs obtained for the ITS2 region, and reads successfully mapped to an ITS2 ASV;
*short*, *ITS*, *LSU*, *long*: ITS2-based ASVs mapped to consensus ASVs for longer regions, where \"short\" and \"long\" denote the full-length short and long amplicons, respectively.
Short amplicons were mapped to full ITS, LSU, and long amplicon ASVs when

@@

@@

(ref:venn-table-asv) Correspondences between ASVs found by different sequencing strategies.
Each row shows the number of ASVs (*ASVs*) shared uniquely by one or more datasets , including the fraction of total ASVs for each dataset (*ASV frac*), the number of reads represented by those ASVs in each dataset (*reads*) and the fraction of total reads for the dataset (*frac*).

@@

(ref:venn-table-otu) Correspondences between 97% OTUs found by different sequencing strategies.

@@

```{r venn, fig.cap = "(ref:venn)", out.height = "30%", fig.show="hold"}
vennplot_ASV <- venn_ASV %>%
  rownames_to_column("set") %>%
  filter(set != "Total") %>%
  mutate(set = map_chr(set,
                       ~ strsplit(., "")[[1]] %>%
                     as.integer() %>%
                     as.logical() %>%
                     magrittr::extract(LETTERS[seq_along(.)], .) %>%
                     paste(collapse = "&"))) %>%
  mutate(
    reads_frac = paste(frac_pb_500, frac_pb_483, frac_is_057, `frac_SH-2257`,
                       sep = "/") %>%
      gsub("/+", "/", .) %>%
      gsub("(^/|/$)", "", .),
    label = paste(ASVs, reads_frac, sep = "\n")
  )

eulerr::venn(deframe(select(vennplot_ASV, set, ASVs)) %>% replace_na(0)) %>%
  plot(
    legend = list(
      labels = c("PacBio Long", "PacBio Short", "Ion Torrent", "Illumina")
    ),
    quantities = vennplot_ASV$label,
    main = "ASV"
  )

vennplot_OTU <- venn_OTU %>%
  rownames_to_column("set") %>%
  filter(set != "Total") %>%
  mutate(set = map_chr(set,
                       ~ strsplit(., "")[[1]] %>%
                     as.integer() %>%
                     as.logical() %>%
                     magrittr::extract(LETTERS[seq_along(.)], .) %>%
                     paste(collapse = "&"))) %>%
  mutate(
    reads_frac = paste(frac_pb_500, frac_pb_483, frac_is_057, `frac_SH-2257`,
                       sep = "/") %>%
      gsub("/+", "/", .) %>%
      gsub("(^/|/$)", "", .),
    label = paste(OTUs, reads_frac, sep = "\n")
  )

eulerr::venn(deframe(select(vennplot_OTU, set, OTUs)) %>% replace_na(0)) %>%
  plot(
    legend = FALSE,
    labels = FALSE,
    quantities = vennplot_OTU$label,
    main = "OTU"
  )

@@

compR2 <- function(x, y, type) {
  x <- plyr::mapvalues(
    x,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine,
      "-",
      datasets$amplicon),
    FALSE
  )
  y <- plyr::mapvalues(
    y,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine,
      "-",
      datasets$amplicon),
    FALSE
  )
  format(filter(comparisons, x == x_var, y == y_var, type == !!type)$r.squared, digits = 2, drop0trailing = FALSE)
}

@@

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in figure \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(venn_ASV[7, c("frac_pb_500", "frac_pb_483", "frac_is_057")]))` of the reads for the long and short PacBio datasets and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(venn_OTU[7, c("frac_pb_500", "frac_pb_483", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, >99% and `r percent(sum(as.numeric(venn_OTU[c("011", "111"), "frac_is_057"])))` of reads in the PacBio and Ion Torrent short-amplicon datasets belonged to OTUs shared between the two datasets.
In contrast, `r percent(venn_ASV["100", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(venn_OTU["100", "frac_pb_500"])` after OTU clustering.

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = `r compR2("is_057", "pb_500", "ASV")` \text{ to } `r compR2("is_057", "pb_483", "OTU")`$).
Comparisons including the Ion Torrent dataset all display a "hook" at low read counts, which is probably due to statistical sampling error when comparing read counts between datasets with very different sequencing depth.
The strongest correlation was for the short amplicon PacBio and Ion Torrent datasets after OTU clustering ($R^2 = `r compR2("is_057", "pb_483", "OTU")`$), which were sequenced from the same amplicon library.

>>>>

(ref:readcounts) Number of Amplicon Sequence Variants (ASV) and reads at different pipeline stages.
*Raw*: raw reads as delivered by sequencing center;
*Trim*: reads with primers and demultiplexing indices removed;
*LSUx*: reads with a positive CM hit for 5.8S, allowing regions to be extracted;
*ITS2*: ASVs obtained for the ITS2 region, and reads successfully mapped to an ITS2 ASV;
*short*, *ITS*, *LSU*, *long*: ITS2-based ASVs mapped to consensus ASVs for longer regions, where \"short\" and \"long\" denote the full-length short and long amplicons, respectively.
Short amplicons were mapped to full ITS, LSU, and long amplicon ASVs when they had a perfect match in the ITS2 region.

@@

readd(allseqs, cache = cache) %>% filter(hash %in% taxa_names(agaricus_reads)) %$%
  set_names(short, hash) %>%
  discard(is.na) %>%
  Biostrings::RNAStringSet() %>%
  Biostrings::writeXStringSet(here("temp/agaricus.fasta"))

@@

(ref:venn-table-asv) Correspondences between ASVs found by different sequencing strategies.
Each row shows the number of ASVs (*ASVs*) shared uniquely by one or more datasets, including the fraction of total ASVs for each dataset (*ASV frac*), the number of reads represented by those ASVs in each dataset (*reads*) and the fraction of total reads for the dataset (*frac*).

@@

(ref:venn-table-otu) Correspondences between 97% OTUs found by different sequencing strategies.
Each row shows the number of OTUs (*OTUs*) shared uniquely by one or more datasets, including the fraction of total OTUs for each dataset (*OTU frac*), the number of reads represented by those OTUs in each dataset (*reads*) and the fraction of total reads for the dataset (*frac*).

@@

```{r venn, fig.cap = "(ref:venn)", out.height = "30%", fig.show="hold"}
vennplot_ASV <- venn_ASV %>%
  rownames_to_column("set") %>%
  filter(set != "Total") %>%
  mutate(set = map_chr(set,
                       ~ strsplit(., "")[[1]] %>%
                     as.integer() %>%
                     as.logical() %>%
                     magrittr::extract(LETTERS[seq_along(.)], .) %>%
                     paste(collapse = "&"))) %>%
  mutate(
    reads_frac = paste(frac_pb_500, frac_pb_483, frac_is_057, `frac_SH-2257`,
                       sep = "/") %>%
      gsub("/+", "/", .) %>%
      gsub("(^/|/$)", "", .),
    label = str_c(ASVs, reads_frac, sep = "\n") %>%
      replace_na("-")
  )

eulerr::venn(deframe(select(vennplot_ASV, set, ASVs)) %>% replace_na(0)) %>%
  plot(
    legend = list(
      labels = c("PacBio Long", "PacBio Short", "Ion Torrent", "Illumina")
    ),
    quantities = vennplot_ASV$label,
    main = "ASV",
    gp = grid::gpar(cex = 0.8)
  )

vennplot_OTU <- venn_OTU %>%
  rownames_to_column("set") %>%
  filter(set != "Total") %>%
  mutate(set = map_chr(set,
                       ~ strsplit(., "")[[1]] %>%
                     as.integer() %>%
                     as.logical() %>%
                     magrittr::extract(LETTERS[seq_along(.)], .) %>%
                     paste(collapse = "&"))) %>%
  mutate(
    reads_frac = paste(frac_pb_500, frac_pb_483, frac_is_057, `frac_SH-2257`,
                       sep = "/") %>%
      gsub("/+", "/", .) %>%
      gsub("(^/|/$)", "", .),
    label = str_c(OTUs, reads_frac, sep = "\n") %>%
      replace_na("-")
  )

eulerr::venn(deframe(select(vennplot_OTU, set, OTUs)) %>% replace_na(0)) %>%
  plot(
    legend = FALSE,
    labels = FALSE,
    quantities = vennplot_OTU$label,
    main = "OTU",
    gp = grid::gpar(cex = 0.6)
  )

@@

compR2 <- function(x, y, type) {
  xamp <- plyr::mapvalues(
    x,
    datasets$seq_run,
    datasets$amplicon,
    FALSE
  )
  yamp <- plyr::mapvalues(
    y,
    datasets$seq_run,
    datasets$amplicon,
    FALSE
  )
  x <- plyr::mapvalues(
    x,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine),
    FALSE
  )
  y <- plyr::mapvalues(
    y,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine),
    FALSE
  )
  formatC(
    filter(
      comparisons,
      x == x_var,
      xamp == x_amplicon,
      y == y_var,
      yamp == y_amplicon,
      type == !!type
    )$r.squared,
    digits = 2,
    format = "f")
}

@@

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in figure \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(venn_ASV["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(venn_OTU["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(venn_OTU[c("0111", "1111"), c("frac_pb_483", "frac_SH-2257", "frac_is_057")], as.numeric))))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(venn_ASV["1000", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(venn_OTU["1000", "frac_pb_500"])` after OTU clustering.

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = compR2("SH-2257", "is_057", "OTU")`$).
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).
